Open-source tools in R for landscape ecology

### 2021-04-26 paper/paper.Rmd
improves glossary

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive \textit{R} Archive Network. A repository for \textit{R} package \\
CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
GDAL & Open source translator library for raster and vector geospatial data formats \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
PROJ & Open source generic coordinate transformation software \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
\textit{R} package & User-created software extension to \textit{R} programming language \\
SDM & Species distribution modelling \\
Simple Features & A set of standards that specify a common storage and access model of geographic features \\
vector & Points, lines, and polygons-based data model \\

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
For more information about *R* package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
*R* allows two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
However, the generic functions are focused on quick visual inspection of the data (Fig. \@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}

The popular plotting package *ggplot2* (Fig. \@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
For more information about *ggplot2* in general, see \cite{Wickham2016}}.
The *tmap* package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) *ggplot2* package, and c) *tmap* package. All three approaches result in similar-looking outputs, but their syntax and capabilities differ. All maps show the total annual precipitation of Switzerland. Data were downloaded with the *raster* package (precipitation data) and *rnaturalearth* package (country borders) package. The code to create the maps can be found in the Appendix.
```{r fig-map, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:capfigmap)"}
include_graphics("data/Figure_2.pdf")
```

@@

\captionsetup{width=2\linewidth}

\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task. The table only includes packages specifically desinged for spatial data or landscape ecology.}

@@

\multirow{3}{*}{Bridge stuff} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\
& RSAGA & Interface between SAGA and R & \cite{Brenning2018} \\
& rgee &  Download data from Google Earth Engine & \cite{Aybar2020} \\

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive \textit{R} Archive Network. A repository for \textit{R} package \\
CRS & Coordinate references system. A coordinate-based local, regional or global system used to locate geographical entities \\
EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
GDAL & Open source translator library for raster and vector spatial data formats \\
GIS & Geographic information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
PROJ & Open-source generic coordinate transformation software \\
raster & Cell grid based spatial data model \\
resolution & Size in meter or degree of one cell \\
\textit{R} package & User-created software extension to \textit{R} programming language \\
SDM & Species distribution modeling \\
Simple Features & A set of standards that specify a common storage and access model of spatial vector data \\
vector & Points, lines, and polygons-based spatial data model \\

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
For more information about *R* package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
*R* allows two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
\textcolor{blue}{Fore more information about **ggplot2** in general see \cite{Wickham2016}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

(ref:capfigmap) Comparison of different options to create maps using a) base plot, b) *ggplot2* package, and c) *tmap* package. All maps show the total annual precipitation of Switzerland. Data were downloaded with the *raster* package (precipitation data) and *rnaturalearth* package (country borders) package. The code to create the maps can be found in the Appendix.
```{r fig-map, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:capfigmap)"}
include_graphics("data/Figure_2.pdf")
```

@@

\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task.}

@@
